Multi-level modeling with Openflexo/\FML

### 2021-10-13 responseLetter.tex

\pagebreak

\Editor{}

\begin{response}{the reviewers have made suggestions for improving the presentation but also raised a number of issues requiring clarification. An overarching theme is the request for a more explicit description of how the multi-level modelling emulation is achieved and what its general advantages and limitations are, including a clarification of what the implications would be of supporting more "levels" than the challenge required.}

%We have been more explicit to explain how we use the FML language at the end of the introduction, and in various figures. TODO.

We have enhanced the presentation of both, Openflexo/FML and our solution to the challenge so that: 1) the techniques we use to \emph{achieve} multi-level capabilities are more explicit; and, 2) the advantages and disadvantages of using them in order to fulfil the challenge requirements become clearer. Notably, we have added many details in the final paragraphs of the introduction section, improved the Section 2, devoted to the description of OpenFlexo, refactored figures 2, 4, 5, 6, 7 and 9 and, added figures 3 and 10.
\end{response}

\begin{response}{We very much welcome a "conventional" response to the challenge but would like to see the respective reviewer questions addressed in a revised version which is explicit as possible about the advantages and the disadvantages of the emulation. You may find the references listed below useful to keep explanations concise by referencing related work (note that reviewer 3’s “Type-Instance pattern” is also known as the “Type Object” pattern). You should also respond to other concerns reviewers raise, either by making changes to the submission or in the form of an authors' response.} 
The previous answer describes the improvements of our revision.
Additionally, the related work section has been modified to show that we are not the first "conventional" response with multi-level emulation, and to improve comparisons and limitations of different approaches ("conventional" or not).

%\todo[inline]{I do not understand this answer. We are saying that other approaches are also unconventional?}
\end{response}

@@

\pagebreak

\Reviewer{\#1}
\begin{response}{In section 2 Technology the authors describe the foundations of the Openflexo approach. Here it would greatly help the reader to see an example how the approach works for traditional modeling languages. For example, how would you realize a small subset of BPMN with the approach? This could briefly be illustrated so as to give the reader an immediate understanding. Especially, the 'metamodel'-agnostic approach seems very interesting but I would like to see how that is used for traditional modeling approaches.}

The Figure \ref{fig:BPMNSubsetExample} (above) shows how to realize a small subset of BPMN with Openflexo/FML (assuming a metamodel of the full specification of BPMN exists). Concretely, we show  how a VirtualModel (SimplifiedBPMN) can filter an existing model (BPMN) and be instantiated. Note that connection with external sources is not required for our solution to the challenge and is only represented in order to illustrate the capabilities of Openflexo/FML. Therefore, in the article we choose not to give this level of details and focus on the left part of the figure.

%\todo[inline]{I've refactored the paragraph above. To be verified.}

@@

\begin{figure}[t]
    \centering
    \includegraphics[width=1.0 \textwidth]{Figures/BPMNSubsetExampleWithExternalConnexion.pdf}
    \caption{An example of BPMN subset built on top of a BPMN/EMF model}
    \label{fig:BPMNSubsetExample}
\end{figure}

%\begin{lstlisting}
%use org.openflexo.technologyadapter.emf.EMFModelSlot as EMF;
%import ["http://www.omg.org/spec/BPMN/20100524/MODEL-XMI"] as BPMN_METAMODEL;
%
%// A simplified BPMN metamodel exposing
%// both concepts BPMNProcess and BPMNTask
%model SimplifiedBPMN {
%
%  // Connexion to a .bpmn file
%  EMFModel bpmnModel 
%    with EMF::EMFModelSlot(
%        metaModel = BPMN_METAMODEL);
%
%  // Abstract concept BPMNElement defining
%  // a name and a link to an eventual BPMNElement
%  // in .bpmn file
%  abstract concept BPMNElement {
%  	EMFObjectIndividual bpmnElement 
%  	    with EMF::EMFObjectIndividualRole (
%			source = bpmnModel
%		);
%    String name values bpmnElement.name;
%  }
%
%  concept BPMNProcess extends BPMNElement {
%
%    // Describe here properties and
%    // behaviours for BPMNProcess concept
%    ...
%
%    concept BPMNTask extends BPMNElement {
%
%        // Describe here properties and
%        // behaviours for BPMNTask concept
%        ...
%
%    }
%
%    // Other core concepts
%  }
%}
%\textbf{}\end{lstlisting}

\end{response}

\begin{response}{Also, at the stage of section 2, the characterization of the FML language should be made clearer. The metamodel in Figure 2 seems mainly conceptual - but at the same time you refer to it as a basis for implementation. Could that be clarified? eg you have the class "Behavior" - which is a concept and I did not understand right away what it is composed of. On the other hand you have for example the class "ModelSlot" which seems like a very technical realization. Also, you note that roles have types but I could not find any types in the metamodel (?). Thus, I propose to include a more detailed version of your metamodel at this stage. Please note that EMISAJ does not have a page limitation, so no worries if that should require more space.} 
We've changed figure 2. It shows \texttt{Type} and the main concepts used by FML. Explanation of figure has been improved. 
\end{response}

\begin{response}{In Figure 5, the base metamodel for representing processes is shown - however it is not linked to the concepts you introduced before. Would it be possible to maybe graphically highlight how it relates to the metamodel in Figure 2? Besides, as multi-level modeling is much different to traditional modeling, you may want to consider using a different notation rather than UML class diagrams - as these are commonly known for traditional modeling approaches. When reading the paper, I actually marked the references to types in yellow in my copy - ie from ProcessType to TaskType and from ActorType to ActorType (on the attribute levels) - this helped me to see more clearly where you have references from attributes to types - but that's just my way of doing it.}

TODO. Faire le lien entre figure 2 et les représentations des figures 3, 4, 5, 6 et 7. Package = VirtualModel, Classes bleues = FlexoConcept, attribut =(~) Get/GetSetProperty, association = FlexoRole, no FlexoBehavior visible. (!! Dans la figure 2 il n'y a pas d'héritage possible visible entre FlexoConcept.) Le niveau orange est l'instantiation / interprétation du FML.
\end{response}

@@

\begin{response}{(3) One of the key benefits of the Type-Instance pattern is that it provides a 'hook' on which semantics can be defined using constraints. There is a reified relationships between a type and its instance and therefore conditions can be defined that establish both static and dynamic semantics. I am confused regaring why this is not a key feature of the semantics for the challenge as it is implemented by the authors. Constraints are provided very sparingly compared to implementation of behaviour in FML. Unfortunately, since the dynamic semantics of FML is not provided (and is bespoke) this is less clear than it might otherwise be for the reader. Perhaps it would be better to use constraints in the context of the Type-Instance pattern to specify semantics? It would be more abstract than the current 'method' style semantics.} 
TODO.
\end{response}

\begin{response}{(4) Section 6 lists a number of benefits of the approach. Unfortunately, the presentation in the section is very general and none of the claims are justified in terms of concrete descriptions and/or examples. This section seems to be very important and could be of great interest to the reader. Perhaps it would be appropriate to reduce some of the rather implementation heavy descriptions earlier in the paper and increase the analysis in this section? In addition section 4.3 notes that MLM is key to the ability of Openflexo to create tooling. Again, this is very interesting, but no concrete details are provided as to how Openflexo uses the level agnotic features of the models to create tools. Can the essence of this be captured for Openflexo and then described in terms of how it facilitiates the tooling arising from the challenge?}

We tried to illustrate subsections of section 6 by giving examples, by referencing figures and sections (subsections 6.2, 6.3, 6.4, 6.6, 6.7 and 6.11 have been updated).

@@

Concerning the fact that "MLM is the key to the ability of Openflexo to create tooling", the reviewer probably read "MLM" instead of "FML" (Federation Modeling Language).
\end{response}

\begin{response}{(5) The paper makes an issue of the model federation approach supported by Openflexo, including the use of model slots. However, after introducing these concepts they do not seem to be important to the rest of the article. Perhaps this is an implementation concern that could be elided as its inclusion does not seem key to the contributions of the article. } 
\todo[inline]{While it is true that model slots and other features of Openflexo/FML are not used in order to build our solution,  we believe including them in our description help readers understand the framework. We therefore choose to keep the description as is.}
\end{response}

### 2021-10-13 10-conclusions.tex

\textcolor{red}{The features of \FML enable a great modelling flexibility which, coupled with an appropriate methodology, allow us to achieve
the multi/level capabilities without dedicated tooling in order to solve problems which require it. We use virtual models to "implement" levels on demand.}
%\noteSM{Say that the flexibility of our approach coupled with a methodology allows us to achieve multi/level capabilities without specific tooling in order to solve problems which require it.}

>>>>

The features of \FML enable a great modelling flexibility which, coupled with an appropriate methodology, allow us to achieve
the multi/level capabilities without dedicated tooling in order to solve problems which require it. We use virtual models to "implement" levels on demand.
%\noteSM{Say that the flexibility of our approach coupled with a methodology allows us to achieve multi/level capabilities without specific tooling in order to solve problems which require it.}

### 2021-10-13 8-discussion.tex

  Behaviors are continuously checked, ensuring the integrity of the models when
  changes occur. However, most of these behaviors have to be written by the
  users. Thus, the integrity of the contents relies on the users when they
  build the metamodels and the models. The section~\ref{sec:model} describes
  how the requirements are captured and how behaviors are implemented with
  \FML. For instance, \texttt{isAuthorizedActor} behavior is needed to enforce
  \textbf{P17} requirement, leading to writing \FML code in \texttt{TaskType}
  and \texttt{Task}. Note that in our approach, the metamodels are built in an
  {\color{red}ad hoc} way together with the models (co-construction), therefore the user
  also validates on an on-demand basis.

@@

Our tooling includes mechanisms to verify the models. First, syntactical
consistency is realized by cardinality checks and by typing. Second, the
constraints the users have written are continuously checked during model
edition and enactment. Thus, a part of the verification relies on the fact that
users add behaviors when they build the metamodels and the models. For
instance, \textbf{S6} requirement is constrained by an invariant implemented in
\texttt{CodeArtifact}. It ensures that \texttt{artifactType} has the right type
and that it implements the \texttt{languages} relation.
\noteJC{updated}

>>>>

  Behaviors are continuously checked, ensuring the integrity of the models when
  changes occur. However, most of these behaviors have to be written by the
  users. Thus, the integrity of the contents relies on the users when they
  build the metamodels and the models. The section~\ref{sec:model} describes
  how the requirements are captured and how behaviors are implemented with
  \FML. For instance, \texttt{isAuthorizedActor} behavior is needed to enforce
  \textbf{P17} requirement, leading to writing \FML code in \texttt{TaskType}
  and \texttt{Task}. Note that in our approach, metamodels are built in an
 ad hoc way together with the models (co-construction), therefore the user
  also validates on an on-demand basis.

@@

Our tooling includes mechanisms to verify the models. First, syntactical
consistency is realized by cardinality checks and by typing. Second, the
constraints the users have written are continuously checked during model
edition and enactment. Thus, a part of the verification relies on the fact that
users add behaviors when they build the metamodels and the models. For
instance, \textbf{S6} requirement is constrained by an invariant implemented in
\texttt{CodeArtifact}. It ensures that \texttt{artifactType} has the right type
and that it implements the \texttt{languages} relation.
%\noteJC{updated}
